Tree size, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
worked on results

*Descriptions of Droughts*

![**Figure 1. ** (a) Time series of peak growing season (May-August) climate conditions and residual chronologies for each species. Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from [@helcoski_growing_2019]. (b) density plots of community- wide resitance values for each drought.](tables_figures/Figure1.png){width=500px}

In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig. 1). 
We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the strongest droughts of the study period (**DETAILS**). Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp. (**Cite Shenanadoah paper, if accepted**). 
Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but *were not among the lowest in terms of PDSI and were not identified as a pointer yeasr.* 
**The droughts differed in timing/duration/etc. .. The year 1966 was preceded by two relatively dry years... 1964 among five driest in terms of May-August [PET-PRE],  1965  also anomalously hot and dry. **

Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered growth reductions of 30% or more (resistance <= 0.7): #% in 1966, #2% un 1977, and #% in 1999. *Some* trees exhibited increased growth: (resistance > 1.0): #% in 1966, #% un 1977, and #% in 1999. Within the context of mixed effects models, there were no significant differences in resistance across drought years (Table 4).

@@

**Results for first main question: what drives the observed tendency for large trees to suffer more during drought?**
H1.0, H1.1, H1.2, H1.3
DBH, height, crown position, and TWI

![Height profile graphs](tables_figures/Figure2.png){width=500px}

>>>>

*Focal droughts and their characteristics*

![**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species. Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from [@helcoski_growing_2019]. Density plots (b) show community- wide resistance values for each drought.](tables_figures/Figure1.png){width=500px}

In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig. 1). 
We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the strongest droughts of the study period (**DETAILS**). Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp. (*Cite Shenanadoah paper, if accepted*). 
Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but *were not among the lowest in terms of PDSI and were not identified as a pointer yeasr.* 
**The droughts differed in timing/duration/etc. .. The year 1966 was preceded by two relatively dry years... 1964 among five driest in terms of May-August [PET-PRE],  1965  also anomalously hot and dry. **

Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% un 1977, and #% in 1999. *Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% un 1977, and #% in 1999. Within the context of mixed effects models, there were no significant differences in resistance across drought years (Table 4).

@@

*Tree size and drought resistance*

Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no signficant effect for 1977 or 1999 individually (Tables 1, 4). The same held true for height as a univariate predictor (*H1.1*; Tables 1, 4). When combined with other predictor variariables in our multivariate models, the top models usually included an effect of height, and its coefficient was consistently negative, as predicted (Tables 1, 5). We note that a non-signficant positive correlation between $H$ and $R$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).
H1.0, H1.1, H1.2, H1.3
DBH, height, crown position, and TWI

![**Figure 2. Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity (RH), (c) air temperature, and (d) biological temperature averaged over each month of the peak growing season (May-August) from YEAR-YEAR. In these plots, heights are slightly offset for visualization purposes. Also shown are (e) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (f) $PLA_{dry}$ and (g) $\pi_{tlp}$. In (f-g), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals. In all plots, the dahsed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}
